Parallelization of Fast $\ell_1$-Minimization for Face Recognition

### 2011-03-16 main.tex
Working on standardizing terminology

\section{Hardware Parallelism} \label{sec:parallelism} 
In this section we discuss the hardware architectures used in this paper. Aside
from the $\ell_1$-min algorithm, our choice of the architectures plays a
significant role in many practical decisions we make in Section
\ref{sec:alignment} and Section \ref{sec:recognition}.  For CPU implementation,
the system makes use of all the cores in as many CPUs are present on one
motherboard.  For GPU implementation, the system makes the best use of a
single many-core Nvidia GPU chip on a PCI card (GPUs are commonly shipped with
a single GPU chip per PCI-express card).  Unless otherwise specified, all
implementations utilize single precision floating point datatypes.  Since on-chip
memory (cache) is often orders of magnitude faster than off-chip memory, the sizes
and speeds of on-chip memory (cache) have big performance implications.  We give
a brief overview of the caches that are available in our target architectures,
and defer discussion of their performance implications to sections
\ref{sec:alignment_implementation_cpu} and
\ref{sec:alignment_implementation_gpu}.

\subsection{CPU Hardware Parallelism} 
The baseline architecture for our experiments is a Linux workstation with two
quad-core Intel Nehalem E5530 processors clocked at 2.4 GHz.  Each processor
has its own memory interface, and is directly interfaced to half of the RAM
installed in the machine.  The amount of RAM installed exceeds the amount used
by the algorithms, and is not an important performance consideration.

@@

There are thus two important levels of parallelism that need to be exploited to
efficiently use a modern CPU: core-level parallelism and vector-level
parallelism. CPU tools for exploiting these two levels of parallelism are
relatively mature.  Heavily optimized Basic Linear Algebra Subprograms (BLAS) are
commercially available that make good use of both levels of CPU parallelism.
For operations that do not fit well with the BLAS API, we 
achieve both vector-level and core-level parallelism through a combination of
the OpenMP API \cite{dagum2002openmp} and the automatic vectorization 
facilities of Intel's C++ compiler (ICC)\cite{dulong1999overview}.
% DW: I'm really not sure these citations are even really relevant;
% our target audience will just google icc openmp and get more up-to-date
% info anyway.
\subsection{GPU Hardware Parallelism}
Compared to CPU architectures, GPU architectures dedicate less of their transistors
to cache, and more of them for arithmetic logic. Our target GPU architecture
is the Nvidia Fermi architecture.  CUDA based GPUs are comprised of several 
streaming multiprocessors (SMP) (currently 14 to 16 for a high end GPU).

@@

\section{Parallelization of the Face Alignment Stage}
\label{sec:alignment}
%{\bf AY: Add an algorithm here describe the alignment block in the pipeline. Explain the algorithm and necessary modification from generic l-1}
This section provides a brief explanation of the iterative alignment stage
of the recognition pipeline.  
The inputs of the algorithm are the test image $f(x,y)$ stacked into an $M$-
dimensional vector $\f$, and the aligned and stacked training images $A_i$ for
each training user $i$. 
The algorithm repeatedly solves a linearized version of the minimization problem
in \eqref{eq:l1min_alignment}.  This involves two computational steps:
the computation of the linearization, and the solution of the minimization problem.

The computation of the linearization is complicated by the need to avoid trivial 
solutions where the algorithm
progressively zooms in on a black pixel of the test image.
This is achieved by minimizing the $\ell_1$ error between the illumination model $Ax$ and the cropped,
resampled, and normalized test image $\tilde\bb$, rather than the un-normalized version as in 
conventional Lucas-Kanade iterative alignment.
This involves a bit of calculus:
\begin{equation}
\frac{\partial}{\partial \tau_p} \frac{\bb}{||\bb||} 
= \frac{1}{||\bb||} \frac{\partial \bb}{\partial \tau_p} - 
\frac{\bb}{||\bb||^3} \bb^T \frac{\partial \bb}{\partial \tau_p}
\end{equation}
If the warping $\tau$ is constrained to be a similarity transform,
it can be paramaterized in such a way that the computation of $\frac{\partial \bb}{\partial \tau_p}$
reduces to several linear vector operations involving pixel coordinates $(u,v)$, $\tau \in \Re^4$,
and the image gradients $\f_x, \f_y \in \Re^M$, the latter of which only has to be computed
once per test image.  The iterative alignment stage is summarized as Algorithm \eqref{alg:iterative_alignment}.
\begin{algorithm}[thb]
\caption{\bf (Iterative Alignment Stage)} \label{alg:iterative_alignment}
\begin{algorithmic}[1]
\begin{small}
\STATE {\bf Input:} Aligned training images $\{A_i \in \Re^{m\times n_i}\}_{i=1}^K$ for $K$ subjects,  a test image
$\f\in\Re^M$ 
%\STATE $\tau^{(0)} \leftarrow \textrm{face detector($\f$)}$. 
\STATE $\f_x, \f_y \in \Re^M \leftarrow \frac{df}{dx}, \frac{df}{dy}$ 
\STATE {\bf for} each gallery subject $i$, 
\STATE \hspace{3mm} {\bf while} not converged $(j=1,2,\ldots)$ {\bf do} 
\STATE \hspace{6mm} $\tilde \bb(\tau) \leftarrow \frac{\bb \circ \tau}{\|\bb \circ \tau\|_2}$;
\STATE \hspace{6mm} $J \leftarrow  \frac{\partial}{\partial \tau} \tilde\bb(\tau)  \bigr|_{\tau^{(j)}} $;
\STATE \hspace{6mm} $\Delta \tau =  \arg\min \; \| \e \|_1  \; \subj \; \tilde \bb + J \Delta \tau = A_i \x + \e.$ 
\STATE \hspace{6mm} $\tau^{(j+1)} \leftarrow \tau^{(j)} + \Delta
\tau$; 
\end{small}
\end{algorithmic}
\end{algorithm}

@@

\subsection{Computational Complexity} Any discussion of the speed of an
algorithm should discuss the comutational complexity of the algorithm.  For the
alignment stage ALM algorithm \eqref{alg:alignment_alm}, this is rather simple.
The Matrix-vector computions are $O(mn)$, and the vector-vector operations are
$O(m)$, in terms of both computation cost and memory-bandwidth.  Since A is
very tall, the computation of $A^\dagger$, the pseudoinverse of $A$, is best
computed via $A^\dagger = (A^TA)^{-1} A^T$.  The complexity of computing $G =
A^T A$ is $O(nnm)$.  $G$ is sufficiently small that it can be inverted via
Gauss-Jordan elimination, which has a complexity of $O(n^3)$.  The computation
of $A$ finishes with a matrix multiplication, with cost $O(n^2 m)$.
Fortunately, $A^\dagger$ only needs to be computed once per instance of the
$\ell_1$ minimization problem.

@@

I summary, the alignment stage essentially two levels of available parallelism:
at the higher level there are multiple alignent problems that are solved
independently, or {\bf problem-level} parallelism.  At a lower level, the first-order linear algebraic operations
exhibit parallelism within image operations, i.e. at a pixel level.  We term
this {\bf pixel-level} parallelism.

\subsection{CPU Implementation} 
\label{sec:alignment_implementation_cpu}

One of the main contributions of this paper is a more efficent mapping of the
two levels of parallelism available in iterative face alignment to the two
levels of concurrency available in atypical multicore CPU architecture.

>>>>

\section{Hardware Concurrency} \label{sec:parallelism}

In this paper we will clearly distinguish between algorithm {\em parallelism},
which is a property of an algorithm, and hardware {\em concurrency}, which is a
property of a hardware architecture.  Parallelism in an algorithm provides the
opportunity to perform computations concurrently on hardware.  Algorithms often
exhibit multiple levels of parallelism, and hardware often provides multiple
levels of concurrency.  In these terms, the primary contribution of this paper
will be to determine the optimal mappings between the parallelism available in
$\ell_1$-Minimization and the concurrency available in multi-core CPU and GPU architectures.

In this section we discuss the levels of concurrency available in the hardware
architectures considered in this paper, as well as other aspects of the
hardware that are important for performance.  In particular, since on-chip
memory (cache) is often orders of magnitude faster than off-chip memory, the
sizes and speeds of on-chip memory (cache) have big performance implications.
We give a brief overview of the caches that are available in our target
architectures, and defer discussion of their performance implications to
sections \ref{sec:alignment_implementation_cpu} and
\ref{sec:alignment_implementation_gpu}.

We will primarily be comparing CPU and GPU architectures on a per-board basis:
for CPU implementation we make use of all the cores in as many CPUs are present
on one motherboard, and for GPU implementation we make use of a single
many-core Nvidia GPU chip on a PCI card. GPUs are commonly shipped with a
single GPU chip per PCI-express card.  Note that most engineering  workstations
are capable of supporting several GPU cards.  Unless otherwise specified, all
implementations utilize single precision floating point datatypes.

\subsection{CPU Hardware Concurrency} 
The main defining characteristics of contemporary multi-core CPU architectures
are that they have two levels of concurrency, relatively large amounts of
on-chip cache, and relatively high clock speeds.

The baseline architecture for our experiments is a Linux workstation with two
quad-core Intel Nehalem E5530 processors clocked at 2.4 GHz.  Each processor
has its own memory interface, and is directly interfaced to half of the RAM
installed in the machine.  The amount of RAM installed exceeds the amount used
by the algorithms, and is not an important performance consideration.

@@

There are thus two important levels of concurrency that need to be exploited to
efficiently use a modern CPU: {\em core-level} concurrency and {\em vector-level}
concurrency. CPU tools for exploiting these two levels of concurrency are
relatively mature.  Heavily optimized Basic Linear Algebra Subprograms (BLAS) are
commercially available that make good use of both levels of CPU concurrency.
For operations that do not fit well with the BLAS API, we 
achieve both vector-level and core-level concurrency through a combination of
the OpenMP API \cite{dagum2002openmp} and the automatic vectorization 
facilities of Intel's C++ compiler (ICC)\cite{dulong1999overview}.
% DW: I'm really not sure these citations are even really relevant;
% our target audience will just google icc openmp and get more up-to-date
% info anyway.
\subsection{GPU Hardware Concurrency}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
on-chip cache, and relatively low clock speeds.

Compared to CPU architectures, GPU architectures dedicate less of their transistors
to cache, and more of them for arithmetic logic. Our target GPU architecture
is the Nvidia Fermi architecture.  CUDA based GPUs are comprised of several 
streaming multiprocessors (SMP) (currently 14 to 16 for a high end GPU).

@@

\section{Parallelism in the Face Alignment Stage}
\label{sec:alignment}
There are essentially two levels of parallelism available for exploitation in
for iterative alignment in the face recognition application.  The primary goal
of this section is to introduce these two levels of parallelism.  Additionally,
in order for some important aspects of our implementations to be clear later,
we need to establish how the iterative alignment stage works in some
detail.

The inputs of the iterative alignment stage are the test image $f(x,y)$ stacked into an $M$-
dimensional vector $\f$, and the aligned and stacked training images $A_i$ for
each training user $i$. 
The algorithm repeatedly solves a linearized version of the minimization problem
in \eqref{eq:l1min_alignment}.  This involves two computational steps:
the computation of the linearization, and the solution of the minimization problem.

The computation of the linearization is complicated by the need to avoid trivial 
solutions where the algorithm
progressively zooms in on a black pixel of the test image.
This is achieved by minimizing the $\ell_1$ error between the illumination model $Ax$ and the cropped,
resampled, and normalized test image $\tilde\bb$, rather than the un-normalized version as in 
conventional Lucas-Kanade iterative alignment.
This involves a bit of calculus:
\begin{equation}
\frac{\partial}{\partial \tau_p} \frac{\bb}{||\bb||} 
= \frac{1}{||\bb||} \frac{\partial \bb}{\partial \tau_p} - 
\frac{\bb}{||\bb||^3} \bb^T \frac{\partial \bb}{\partial \tau_p}
\end{equation}
If the warping $\tau$ is constrained to be a similarity transform,
it can be paramaterized in such a way that the computation of $\frac{\partial \bb}{\partial \tau_p}$
reduces to several linear vector operations involving pixel coordinates $(u,v)$, $\tau \in \Re^4$,
and the image gradients $\f_x, \f_y \in \Re^M$, the latter of which only has to be computed
once per test image.  The per-user iterative alignment algorithm is summarized as Algorithm \eqref{alg:iterative_alignment}.
\begin{algorithm}[thb]
\caption{\bf (Per-User Iterative Alignment)} \label{alg:iterative_alignment}
\begin{algorithmic}[1]
\begin{small}
\STATE {\bf Input:} Aligned training images $\{A_i \in \Re^{m\times n}\}$ for gallery subject $i$, the test image
$\f\in\Re^M$ 
%\STATE $\tau^{(0)} \leftarrow \textrm{face detector($\f$)}$. 
%\STATE $\f_x, \f_y \in \Re^M \leftarrow \frac{df}{dx}, \frac{df}{dy}$ 
%\STATE {\bf for} each gallery subject $i$, 
\STATE \hspace{0mm} {\bf while} not converged $(j=1,2,\ldots)$ {\bf do} 
\STATE \hspace{3mm} $\tilde \bb(\tau) \leftarrow \frac{\bb \circ \tau}{\|\bb \circ \tau\|_2}$;
\STATE \hspace{3mm} $J \leftarrow  \frac{\partial}{\partial \tau} \tilde\bb(\tau)  \bigr|_{\tau^{(j)}} $;
\STATE \hspace{3mm} $\Delta \tau =  \arg\min \; \| \e \|_1  \; \subj \; \tilde \bb + J \Delta \tau = A_i \x + \e.$ 
\STATE \hspace{3mm} $\tau^{(j+1)} \leftarrow \tau^{(j)} + \Delta
\tau$; 
\end{small}
\end{algorithmic}
\end{algorithm}

@@

\subsection{Computational Complexity and Parallelism} Any discussion of the speed of an
algorithm should discuss the comutational complexity of the algorithm.  For the
alignment stage ALM algorithm \eqref{alg:alignment_alm}, this is rather simple.
The Matrix-vector computions are $O(mn)$, and the vector-vector operations are
$O(m)$, in terms of both computation cost and memory-bandwidth.  Since A is
very tall, the computation of $A^\dagger$, the pseudoinverse of $A$, is best
computed via $A^\dagger = (A^TA)^{-1} A^T$.  The complexity of computing $G =
A^T A$ is $O(nnm)$.  $G$ is sufficiently small that it can be inverted via
Gauss-Jordan elimination, which has a complexity of $O(n^3)$.  The computation
of $A$ finishes with a matrix multiplication, with cost $O(n^2 m)$.
Fortunately, $A^\dagger$ only needs to be computed once per instance of the
$\ell_1$ minimization problem.

@@

In summary, the alignment stage essentially two levels of available
parallelism: at the higher level there are multiple iterative alignent problems
that are solved independently, or {\bf problem-level} parallelism.  At a lower
level, the first-order linear algebraic operations exhibit parallelism within
image operations, i.e. at a pixel level.  We term this {\bf pixel-level}
parallelism.

\subsection{CPU Implementation} 
\label{sec:alignment_implementation_cpu}
One of the main contributions of this paper is a more efficent mapping of the
two levels of parallelism available in iterative face alignment to the two
levels of concurrency available in a typical multicore CPU architecture.
